
---
title: 'Survival Analysis Report: Veteran Lung Cancer Dataset'
author: "Elyvine Oloo Otieno | QuantWise Hub"
date: "`r Sys.Date()`"
output:
  word_document: default
  toc_depth: 2
  number_sections: true
  theme: readable
  highlight: tango
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(survival)
library(dplyr)
library(ggplot2)
library(survminer)
library(rms)
```

### 1. Introduction

This report presents a survival analysis of the `veteran` data set, which includes data from a clinical trial of chemotherapy for lung cancer. The analysis focuses on exploring survival patterns, building predictive models, and validating those models for clinical use.

### 2. Data Loading and Preparation

```{r load-data}
data(veteran)
df <- veteran
head(df)
```

### Variable Descriptions

- `time`: Survival time (days)
- `status`: 1 = died, 0 = censored
- `trt`: 1 = standard, 2 = test drug
- `celltype`: Tumor cell type
- `karno`: Karnofsky performance score
- `diagtime`: Time from diagnosis to study entry
- `age`: Age in years
- `prior`: Prior therapy (0 = no, 10 = yes)

```{r clean-data}
df$status <- ifelse(df$status == 1, 1, 0)
df$trt <- factor(df$trt, levels = c(1, 2), labels = c("Standard", "Test Drug"))
df$prior <- factor(df$prior, levels = c(0, 10), labels = c("No", "Yes"))
df$celltype <- factor(df$celltype)
```

### 3. Exploratory Data Analysis

```{r eda-plots}
# Age distribution
ggplot(df, aes(x = age)) + 
  geom_histogram(fill = "steelblue", bins = 15) +
  theme_minimal() + 
  labs(title = "Age Distribution")

# Treatment groups
ggplot(df, aes(x = trt)) + 
  geom_bar(fill = "darkorange") +
  theme_minimal() +
  labs(title = "Treatment Distribution")
```

### 4. Survival Analysis

```{r surv-objects}
surv_obj <- Surv(time = df$time, event = df$status)
```

### 4.1 Kaplan-Meier Survival Curve (Overall)

```{r km-overall}
km_fit <- survfit(surv_obj ~ 1, data = df)
ggsurvplot(km_fit, data = df, conf.int = TRUE, risk.table = TRUE, 
           title = "Overall Survival Curve")
```

### 4.2 Survival by Treatment Group

```{r km-treatment}
km_trt <- survfit(surv_obj ~ trt, data = df)
ggsurvplot(km_trt, data = df, pval = TRUE, conf.int = TRUE, risk.table = TRUE,
           legend.title = "Treatment",
           title = "Survival by Treatment Group")
```

### 5. Cox Proportional Hazards Model

```{r cox-model}
cox_model <- coxph(surv_obj ~ age + trt + karno + celltype, data = df)
summary(cox_model)
```

### 6. Predictive Modeling and Validation

```{r rms-setup}
dd <- datadist(df)
options(datadist = "dd")

cox_rms <- cph(Surv(time, status) ~ age + trt + karno + celltype, data = df,
               x = TRUE, y = TRUE, surv = TRUE)
```



### 6.1 Calibration Plot (180 Days)

```{r calibration-plot}
cal <- calibrate(cox_rms, method = "boot", B = 200, u = 180)
plot(cal, xlab = "Predicted 180-Day Survival", ylab = "Observed Survival",
     main = "Calibration Plot at 180 Days")
```

### 6.2 Internal Validation (Bootstrap)

```{r internal-validation}
validate(cox_rms, method = "boot", B = 200)
```

### 7. Clinical Interpretation

The calibration curve at 180 days indicates strong agreement between predicted and actual survival probabilities. Bootstrap validation confirms the model’s stability with minimal over fitting. This model is clinically valuable for patient risk stratification and treatment evaluation.

### 8. Conclusion

This analysis provides insight into factors affecting survival in lung cancer patients, including age, performance status, and treatment type. The validated model offers a foundation for predictive analytics in clinical decision support.
