---
title: "Analyzing Regional Trends in Influenza-Like Illness (ILI) in Kenya (2023–2024): A Quantitative Epidemiology Case Study"
author: "Elyvine Oloo Otieno"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
  pdf_document:
    toc: true
always_allow_html: true
---

Required libraries:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)   # Core packages for data manipulation and visualization.
library(ggplot2)     # Specifically used for creating data visualizations
library(dplyr)       # Provides functions for data manipulation 
library(kableExtra)  # Enhances knitr tables with styling options
library(ggpubr)      # Additional plotting tools.
library(broom)       # Tabular representation of ANOVA Results
```

Data Import and Preparation:

```{r load-data}
data <- read.csv("C:/Users/R/Desktop/Data Analysis/R/data/Epi_Task_Data.csv", stringsAsFactors=TRUE)
head(data)
```

1(a). Mean ILI Percentage per County per Year == 2023:

```{r}
mean_ili <- data %>%
  filter(year == 2023) %>%
  group_by(county, year) %>%
  summarise(mean_ili = mean(ili_percentage, na.rm = TRUE)) %>%
  arrange(desc(mean_ili))

kable(mean_ili, caption = "Mean ILI Percentage per County in 2023") %>%
  kable_styling(full_width = FALSE)

```



1(b). Mean ILI Percentage per County per Year == 2024:

```{r mean-ili}
mean_ili <- data %>%
  filter(year == 2024) %>%
  group_by(county, year) %>%
  summarise(mean_ili = mean(ili_percentage, na.rm = TRUE)) %>%
  arrange(desc(mean_ili))

kable(mean_ili, caption = "Mean ILI Percentage per County in 2024") %>%
  kable_styling(full_width = FALSE)

```

2. Weekly ILI Trends & Peak Weeks:

```{r ili-trend}
ggplot(data, aes(x = epi_week, y = ili_percentage, color = county)) +
  stat_summary(fun = mean, geom = "line") +
  labs(title = "Weekly ILI Trends by County (2023-2024)",
       x = "Epidemiological Week",
       y = "Mean ILI Percentage") +
  theme_minimal()
```

3. Incidence Rates per 100,000 (Nairobi, Kisumu, Mombasa):

```{r incidence-rate}
selected <- data %>%
  filter(county %in% c("Nairobi", "Kisumu", "Mombasa")) %>%
  mutate(incidence_rate = (ili_percentage / 100) * 100000)

incidence_summary <- selected %>%
  group_by(county) %>%
  summarise(mean_incidence = mean(incidence_rate, na.rm = TRUE))

kable(incidence_summary, caption = "Mean Incidence Rate per 100,000 Population") %>%
  kable_styling(full_width = F)
```

 4. Comparing ILI Percentages (ANOVA):

```{r anova}
anova_data <- data %>%
  filter(county %in% c("Nairobi", "Kisumu", "Mombasa"))

anova_result <- aov(ili_percentage ~ county, data = anova_data)

# Tidy ANOVA summary
anova_table <- tidy(anova_result)

# Display as a styled table
kable(anova_table, caption = "ANOVA Results: ILI Percentage by County") %>%
  kable_styling(full_width = FALSE)

```

5. Summary of Findings

- Nairobi and Mombasa have generally higher mean ILI percentages.
- Incidence rates indicate a higher ILI burden in Nairobi County.
- Weekly trends show seasonal ILI peaks between weeks 4 and 5.
- ANOVA test shows no significant diffrence differences in ILI proportions among counties.
